---
title: "VCF Quality Control & Relatedness Tutorial"
subtitle: "SNP-based QC, diversity, relatedness, and breeding-group selection"
author: "Your Name"
date: today
format:
  html:
    code-fold: show
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## Overview

This tutorial walks through a typical population-genomics QC and relatedness
workflow starting from a VCF file:

1. Package setup
2. General SNP-level quality control (QUAL, depth, missingness)
3. Basic diversity statistics (Ho, He, Fis)
4. Individual heterozygosity and inbreeding
5. Pairwise relatedness and kinship
6. Network visualization of relatedness
7. Selecting low-relatedness breeding groups with `swingeR`

Adjust the file paths in @sec-setup to point at your own VCF and metadata
files before rendering.

## Setup {#sec-setup}

### Install and load packages

```{r}
#| label: packages
#| eval: false

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

pkgs <- c("vcfR", "ggplot2", "dplyr", "tidyr", "patchwork", "reshape2",
          "tidyverse", "vegan", "dartR.base", "dartR.popgen", "adegenet",
          "SNPRelate", "gdsfmt", "hierfstat", "visNetwork")

for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) {
    if (p %in% c("vcfR", "SNPRelate", "gdsfmt")) {
      BiocManager::install(p)
    } else {
      install.packages(p)
    }
  }
}

# swingeR: install from the local .tar.gz (see the swingeR package repo/README)
install.packages("swingeR_0.2.0.tar.gz", repos = NULL, type = "source")
```

```{r}
#| label: library-load

library(swingeR)
library(vcfR)
library(ggplot2)
library(dplyr)
library(dartR.base)
library(dartR.popgen)
library(tidyverse)
library(vegan)
library(tidyr)
library(patchwork)
library(reshape2)
library(SNPRelate)
library(hierfstat)
library(visNetwork)
```

### File paths

Set these once, and every chunk below will use them.

```{r}
#| label: paths

vcffile  <- "SPBRsimulated.vcf.gz"
gdsfile  <- "simulated.gds"
metafile <- "SPBRmeta.txt"
```

### Load the VCF and metadata

```{r}
#| label: load-data

vcf  <- read.vcfR(vcffile)
meta <- read.table(metafile, header = TRUE, sep = "\t")
```

## General QC checks

### QUAL score distribution

```{r}
#| label: qual-dist

ourSNPs <- vcf2df(vcf)

qd_df <- data.frame(QD = ourSNPs$QUAL_num[is.finite(ourSNPs$QUAL_num)])

ggplot(qd_df, aes(x = QD)) +
  geom_histogram(bins = 100, fill = "#8172B2", colour = "white") +
  xlim(0, 5000) +
  labs(x = "QUAL", y = "Count") +
  theme_bw(base_size = 12)
```

Quality scaled by depth of coverage (QUAL / DP) is often a more informative
filter than raw QUAL alone:

```{r}
#| label: qual-per-dp

ourSNPs$QUAL_per_dp <- ourSNPs$QUAL_num / ourSNPs$DP_raw

qd_df <- data.frame(QD = ourSNPs$QUAL_per_dp[is.finite(ourSNPs$QUAL_per_dp)])

ggplot(qd_df, aes(x = QD)) +
  geom_histogram(bins = 100, fill = "#8172B2", colour = "white") +
  geom_vline(xintercept = 0.2, linetype = "dashed", colour = "red") +
  xlim(0, 3) +
  labs(x = "QUAL / DP (quality / depth of coverage)", y = "Count") +
  theme_bw(base_size = 12)
```

### Read depth (DP) distribution

```{r}
#| label: dp-dist

dp_df <- data.frame(DP = ourSNPs$DP_raw)

ggplot(dp_df, aes(x = DP)) +
  geom_histogram(bins = 100, fill = "#8172B2", colour = "white") +
  labs(title = "Read Depth (DP) Distribution", x = "Depth", y = "Count") +
  theme_bw(base_size = 13)
```

Depth per genotype, trimmed to the 99th percentile, with mean +/- 3 SD
thresholds marked:

```{r}
#| label: dp-per-genotype

dp_per_genotype <- ourSNPs$DP_raw / ourSNPs$n_genotypes
dp_df <- data.frame(DP = dp_per_genotype)
dp_df <- dp_df[!is.na(dp_df$DP) & dp_df$DP < quantile(dp_df$DP, 0.99, na.rm = TRUE), ,
               drop = FALSE]

dp_mean    <- mean(dp_df$DP, na.rm = TRUE)
dp_mean2sd <- dp_mean + 3 * sd(dp_df$DP, na.rm = TRUE)

ggplot(dp_df, aes(x = DP)) +
  geom_histogram(bins = 100, fill = "#8172B2", colour = "white") +
  geom_vline(xintercept = c(6, dp_mean, 85), linetype = "dashed",
             colour = c("red", "orange", "red")) +
  labs(title = "Read Depth (DP) Distribution", x = "Depth", y = "Count") +
  theme_bw(base_size = 13)
```

### Missing data

Per sample:

```{r}
#| label: missing-per-sample

ourInd <- vcf2inddf(vcf)

miss_df <- data.frame(Sample = ourInd$ID, Missing_pct = ourInd$miss_pct)

ggplot(miss_df, aes(x = reorder(Sample, -Missing_pct), y = Missing_pct)) +
  geom_col(fill = "#8172B2") +
  geom_hline(yintercept = 20, linetype = "dashed", colour = "red") +
  labs(title = "Missing Genotypes per Sample", x = "Sample", y = "Missing (%)") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

Per locus:

```{r}
#| label: missing-per-locus

miss_var_df <- data.frame(Miss_pct = ourSNPs$missing)

ggplot(miss_var_df, aes(x = Miss_pct)) +
  geom_histogram(bins = 50, fill = "#CCB974", colour = "white") +
  geom_vline(xintercept = 20, linetype = "dashed", colour = "red") +
  labs(title = "Missing Data per Variant", x = "Missing Genotypes (%)",
       y = "Number of Variants") +
  theme_bw(base_size = 13)
```

### Per-chromosome variant count

```{r}
#| label: per-chrom

chr_df <- ourSNPs %>%
  count(CHROM) %>%
  arrange(as.numeric(gsub("scaffold_", "", CHROM))) %>%
  mutate(CHROM = factor(CHROM, levels = CHROM))

ggplot(chr_df, aes(x = CHROM, y = n, fill = CHROM)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Variants per Chromosome", x = "Chromosome", y = "Count") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Basic diversity statistics

```{r}
#| label: basic-diversity

gi <- vcfR2genind(vcf)
pop(gi) <- as.factor(rep("pop1", nInd(gi)))

stats <- basic.stats(gi)
stats$overall[c("Ho", "Hs", "Fis")]
```

### Per-locus heterozygosity and Fis

```{r}
#| label: perloc-ho

ggplot(stats$perloc, aes(x = Ho)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "red") +
  labs(title = "Observed heterozygosity per locus", x = "Ho",
       y = "Number of Variants") +
  theme_bw(base_size = 13)
```

```{r}
#| label: perloc-he

ggplot(stats$perloc, aes(x = Hs)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 0.75, linetype = "dashed", colour = "red") +
  labs(title = "Expected heterozygosity per locus", x = "He",
       y = "Number of Variants") +
  theme_bw(base_size = 13)
```

```{r}
#| label: perloc-fis

ggplot(stats$perloc, aes(x = Fis)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "red") +
  labs(title = "Inbreeding index per locus (Fis)", x = "Fis",
       y = "Number of Variants") +
  theme_bw(base_size = 13)
```

## Individual heterozygosity

```{r}
#| label: ind-het

df_ihe <- ourInd[order(ourInd$Ho), ]
df_ihe$ID <- factor(df_ihe$ID, levels = df_ihe$ID)

mean_ho <- mean(df_ihe$Ho, na.rm = TRUE)

ggplot(df_ihe, aes(x = ID, y = Ho)) +
  geom_col(fill = "steelblue", colour = "white") +
  geom_hline(yintercept = mean_ho, colour = "tomato", linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = nrow(df_ihe), y = mean_ho * 1.05,
           label = paste("Mean =", round(mean_ho, 3)),
           colour = "tomato", hjust = 1, size = 3.5) +
  labs(title = "Observed heterozygosity per individual", x = "Sample",
       y = "Ho (proportion of heterozygous loci)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

## Individual inbreeding coefficient (GDS/SNPRelate)

```{r}
#| label: vcf2gds

showfile.gds(closeall = TRUE)
genofile <- vcf2gds(vcffile, gdsfile)
snpgdsSummary(gdsfile)
```

```{r}
#| label: filter-chroms

snp_id <- read.gdsn(index.gdsn(genofile, "snp.id"))
chrom  <- read.gdsn(index.gdsn(genofile, "snp.chromosome"))
chrom_keep <- paste0("scaffold_", 1:24)
snps_chrom <- snp_id[chrom %in% chrom_keep]
cat("SNPs on first 24 scaffolds/chromosomes:", length(snps_chrom), "\n")
```

```{r}
#| label: individual-F

ibd <- snpgdsIndInb(genofile, snp.id = snps_chrom, autosome.only = FALSE)
ibd_df <- data.frame(ID = ibd$sample.id, F = ibd$inbreeding)
ibd_df <- ibd_df[order(ibd_df$F), ]

ggplot(ibd_df, aes(x = reorder(ID, F), y = F)) +
  geom_col(fill = "steelblue", colour = "white") +
  geom_hline(yintercept = 0, colour = "tomato", linetype = "dashed") +
  labs(title = "Individual inbreeding coefficient (F)", x = "Sample", y = "F") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

## Pairwise relatedness and kinship

Beta-based relatedness:

```{r}
#| label: relatedness-beta

relBet <- snpgdsIndivBeta(genofile, snp.id = snp_id, autosome.only = FALSE)
rel_matrix <- relBet$beta
rownames(rel_matrix) <- relBet$sample.id
colnames(rel_matrix) <- relBet$sample.id

rel_df <- as.data.frame(as.table(rel_matrix))
colnames(rel_df) <- c("from", "to", "kinship")
rel_df <- rel_df %>% filter(as.character(from) != as.character(to))
```

KING-robust kinship:

```{r}
#| label: relatedness-king

king <- snpgdsIBDKING(genofile, snp.id = snps_chrom, type = "KING-robust",
                       autosome.only = FALSE, verbose = TRUE)
kinship_matrix <- king$kinship
rownames(kinship_matrix) <- king$sample.id
colnames(kinship_matrix) <- king$sample.id

kinship_df <- as.data.frame(as.table(kinship_matrix))
colnames(kinship_df) <- c("from", "to", "kinship")
kinship_df <- kinship_df %>% filter(as.character(from) != as.character(to))

cat("Number of pairwise comparisons:", nrow(kinship_df), "\n")
summary(kinship_df$kinship)
```

### Visualizing the kinship distribution

```{r}
#| label: kinship-plot

kinship_labels <- data.frame(
  x = c(0, 0.125, 0.25, 0.5),
  label = c("Unrelated", "3rd degree", "2nd degree", "1st degree")
)

ggplot(kinship_df, aes(x = kinship)) +
  geom_histogram(bins = 100, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = kinship_labels$x, linetype = "dashed",
             colour = "red", linewidth = 0.8) +
  geom_text(data = kinship_labels, aes(x = x, y = Inf, label = label),
            inherit.aes = FALSE, colour = "tomato", hjust = 1.2, vjust = 1.5,
            angle = 90, size = 4) +
  labs(title = "Pairwise Kinship", x = "Kinship", y = "Frequency") +
  theme_bw(base_size = 13)
```

## Network visualization of relatedness

```{r}
#| label: network

net <- networkY(kinship_df, meta, td = 0.1, method = "fixed")
net
```

```{r}
#| label: network-save
#| eval: false

visSave(net, "relatedness_network.html")
```

## Selecting low-relatedness breeding groups

Combine metadata with the individual inbreeding estimates, then hand the
kinship data frame and combined metadata to `swingeR()` from
the `swingeR` package.

```{r}
#| label: combine-metadata

combined_df <- merge(meta, ibd_df, by = "ID")
combined_df
```

```{r}
#| label: breeding-groups

results <- swingeR(
  kinship_df = kinship_df,
  info_df    = combined_df,
  n_groups   = 2,
  n_females  = 2,
  n_males    = 3
)
print_breeding_groups(results)
```

With tighter thresholds:

```{r}
#| label: breeding-groups-strict

results_strict <- swingeR(
  kinship_df = kinship_df,
  info_df    = combined_df,
  n_groups   = 2,
  n_females  = 2,
  n_males    = 3,
  max_ir_f   = 0.0,
  max_ir_m   = 0.0,
  max_pair_rel_f    = -0.01,
  max_pair_rel_m    = -0.01,
  max_group_avg_rel = -0.02
)
print_breeding_groups(results_strict)
```
